Applying Renormalization Group Theory to the Square Well Liquid

### 2014-05-20 papers/thesis-roth/project.tex
Project report writing Tested on elliot. Took 0:32.16.

% pair potential
\newcommand{\1}{\ensuremath{\textbf{r}_1}}
\newcommand{\2}{\ensuremath{\textbf{r}_2}}

@@

%% \fixme{The following comes from Forte~\cite{Forte11} and doesn't make a whole lot of sense to me}

%% The free energy for each iteration is given by
%% \begin{equation}
%%   f_i(T,n) = f_{i-1}(T,n) + \delta f_i(T,n)\ .
%% \end{equation}
%% The total free is obtained when one takes the
%% iterations to infinity and adds in the attractive component:
%% \begin{equation}
%%   f(T,n) = \lim_{i \to \infty} f_i(T,n) + \fattr\ .
%% \end{equation}
%% The initial free energy density, $f_0$ depends on the liquid model of study. For this work, I use the square well model:
%% \begin{equation}
%%   f_0(T,n) = \fid + \fhs + \frac{a_2^{SW}(T,n)}{\kT}\ ,
%% \end{equation}
%% where $a_2^{SW}(T,n)$ is the second-order term in a high-temperature expansion around $1/\kT$. (See ref~\cite{Gil-Villegas97}.)
%% \fixme{End of Forte confusion}

@@

In Gil-Villegas' paper, HS and Dispersion are wrapped up into the
\textit{monomer} contribution, expressed in terms of energy densities
as:
\begin{align}
  f_\text{mono} = f_\text{HS} - \frac{\alpha^\text{VDW}n}{kT}
\end{align}
with $\alpha^\text{VDW}$ given by
\begin{align}
  \alpha^\text{VDW} &= 2\pi\epsilon\int_\sigma^\infty r^2\phi(r)\,dr \\
  \intertext{using reduced units of $x = r/\sigma$, we have}
  &= 2\pi\sigma^3\int_1^\infty x^2\phi(x)\,dx \\
  &= 3b^\text{VDW}\epsilon\int_1^\infty x^2\phi(x)\,dx
\end{align}
where $b^\text{VDW}$ is the van der Waals size parameter. It corresponds to the volume excluded by two spherical particles of volume $b$: $b^\text{VDW} = 4b = 4\left(\pi\sigma^3/6\right)$.

@@

The pair potential $v(\1,\2)$ describes the interaction between two hard spheres. For the square-well model, the pair potential is
\begin{align}
  v_\text{SW}(\1,\2) &=
    \begin{cases}
      \infty & r < \sigma \\
      -\epsilon & \sigma < r < \lambda\sigma \\
      0 & \lambda\sigma < r
    \end{cases}
\end{align}

@@

Given the \emph{configuration integral} (the potenial energy component of the partition function)
\begin{align}
  Z_N &= \int d\mathbf{r}^N\, e^{-\beta V_N}
\end{align}
and the non-ideal part of the free energy (i.e.~the excess free energy) is
\begin{align}
  F &= -\kT\ln\left( \frac{Z_N}{V^N} \right)
\end{align}
where $V^N$ is the total volume \fixme{I think?}, then the derivative of the excess free energy with respect to the coupling parameter $\lambda$ is
\begin{align}
  \frac{\partial F}{\partial\lambda} &= -\kT\left( \frac{1}{Z_N}\frac{\partial Z_N}{\partial\lambda} \right) \\
  &= -\kT\frac{1}{Z_N}\int d\mathbf{r}^N\, e^{-\beta V_N}(-\beta)\frac{\partial V_N}{\partial\lambda} \\
  \hookrightarrow \beta\frac{\partial F}{\partial\lambda} &= \beta\left\langle V_N^{'}(\lambda) \right\rangle_\lambda \\
  \hookrightarrow \beta F &= \beta F_0 + \beta\int_0^1 d\lambda\, \left\langle V_N^{'}(\lambda) \right\rangle_\lambda
\end{align}
where $\left\langle V_N^{'}(\lambda) \right\rangle_\lambda$ is the canonical ensemble average. We can take a series expansion of this average:
\begin{align}\begin{split} % too long; use split environment, and align to a space (&{})
  \left\langle V_N^{'}(\lambda) \right\rangle_\lambda = &{} \left\langle V_N^{'}(\lambda)\right\rangle_{\lambda = 0} + (\lambda)\frac{\partial}{\partial\lambda}\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda}\bigg|_{\lambda = 0} \\ &{} + \mathcal{O}(\lambda^2)
\end{split}\end{align}
The derivative term is
\begin{align}
  \frac{\partial}{\partial\lambda}\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda} &= \left\langle V_N^{''} \right\rangle_\lambda - \beta\sigma_\lambda^2
\end{align}
where
\begin{align}
  \sigma_\lambda^2 &\equiv \left\langle \left[ V_N^{'}(\lambda) \right]^2 \right\rangle_\lambda - \left\langle V_N^{'}(\lambda) \right\rangle^2_\lambda
\end{align}
Putting this together, we have
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta(1 - 0)\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}(1 - 0)^2\left(\left\langle V_N^{''}(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right)
  \end{split} \\
  \begin{split}
    =&{}  \beta F_0 + \beta\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}\left(\left\langle V_N^{''}(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right) \label{eqn:complicated-HTE}
  \end{split}
\end{align}
Now, we can define the total perturbation energy as
\begin{align}
  W_N &= \sum_{i=1}^N\sum_{j>i}^N w_\lambda(\mathbf{r}_i,\mathbf{r}_j)
\end{align}
Then, following from equation~\ref{eqn:small-perturbation},
\begin{align}
  V_N^{'} &= \frac{\partial}{\partial\lambda}\left(\sum_{i=1}^N\sum_{j>i}^N \lambda w(\mathbf{r}_i,\mathbf{r}_j)\right) \nonumber \\
  &= W_N \\
  \therefore V_N^{''} &= 0
\end{align}
With this in mind, we can simplify equation~\ref{eqn:complicated-HTE} to the \emph{high-temperature expansion}, first derived by Zwanzig:\cite{Zwanzig54}
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta\left\langle W_N \right\rangle_0  \\ &{} - \frac{1}{2}\beta^2\left( \left\langle W_N^2 \right\rangle_0 - \left\langle W_N \right\rangle_0^2 \right) + \mathcal{O}(\beta^3)
  \end{split}
\end{align}
Zwanzig further showed that the $n^{th}$ term could be written in terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle W_N\right\rangle_0 \right]^\nu \right\rangle_0$ with $\nu \leq n$.

Writing the statistical averages in terms of the pair density $n_\lambda^{(2)}(\1,\2)$ (i.e. the probability of having a sphere at \1 and \2 \fixme{make this more clear?}) gives us
\begin{align}
  \frac{\beta F}{N} &= \frac{\beta F_0}{N} + \frac{\beta}{2N}\int_0^1 d\lambda\, \iint d\1 d\2\, n_\lambda^{(2)}(\1,\2)w(\1,\2) \label{eqn:Henderson5-2-12}
\end{align}
The pair density itself can be expanded as
\begin{align}
  n_\lambda^{(2)}(\1,\2) &= n_0^{(2)}(\1,\2) + \lambda\frac{\partial n_\lambda^{(2)}(\1,\2)}{\partial\lambda}\bigg|_{\lambda = 0} + \mathcal{O}(\lambda^2)
\end{align}
Applying this to the second term in equation~\ref{eqn:Henderson5-2-12}, we have
\begin{align}
  \frac{\beta F_1}{N} &= \frac{\beta}{2N}\iint d\1 d\2\, n_0^{(2)}(\1,\2)w(\1,\2) \\
  &= \frac{\beta n}{2}\int d\mathbf{r}_{12}\, g_0(\1,\2)
\end{align}

@@

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=\columnwidth]{figs/SW-RG-compare-phi-lowT}
  \end{center}
  \caption{Free energy density for RG at low temperature}
  \label{fig:SW-RG-compare-lowT}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width=\columnwidth]{figs/SW-RG-compare-phi-highT}
  \end{center}
  \caption{Free energy density for RG near critical temp}
  \label{fig:SW-RG-compare-highT}
\end{figure}

>>>>

% pair potential
\newcommand{\1}{\ensuremath{\textbf{r}_1}}
\newcommand{\2}{\ensuremath{\textbf{r}_2}}
\newcommand{\3}{\ensuremath{\textbf{r}_3}}
\newcommand{\4}{\ensuremath{\textbf{r}_4}}

@@

Forte gives $\delta f_i(T,n)$ as
\begin{align}
  \delta f_i(T,n) &= -\frac{\kT}{V_D}\ln\left[ \frac{Z_D(T,n)}{Z_D^*(T,n} \right]
\end{align}
where
\begin{align}
  Z_D(T,n) &= \int dx\, \exp\left[ -\frac{V_D}{\kT}\left( \bar{f}_D(T,n) + \bar{u}_D(T,n \right) \right] \\
  Z_D^*(T,n) &= \int dx\, \exp\left[ -\frac{V_D}{\kT}\bar{f}_D(T,n) \right]
\end{align}
and
\begin{align}
  \bar{f}_D(T,n) &= \frac{f_{i-1}(T,n+x) + f_{i-1}(T,n-x)}{2} - f_{i-1}(T,n) \\
  \bar{u}_D(T,n) &= \frac{u(T,n+x,\lambda_D) + u(T,n-x,\lambda_D)}{2} - u(T,n,\lambda_D)
\end{align}
Finally, $u(T,n)$ is given by
\begin{align}
  \begin{split}
  u(T,n) = &{} g^\text{HS}(n_{eff};\sigma)\alpha - \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)\alpha\omega^2 \\ &{} + \left( \frac{4\pi^2}{2^{2i+1}L^2} \right)^2\alpha\gamma
  \end{split}
\end{align}
with
\begin{align}
  \alpha &= \frac{2}{3}\pi\epsilon\sigma^3(\lambda^3 - 1) \\
  \omega^2 &= \frac{1}{5}\sigma^2\frac{\lambda^5 - 1}{\lambda^3 - 1} \\
  \gamma &= \frac{1}{70}\sigma^4\frac{\lambda^7 - 1}{\lambda^3 - 1}
\end{align}

@@

In Gil-Villegas' paper,\cite{gil-villegas97} HS and Dispersion are wrapped up into the
\textit{monomer} contribution, expressed in terms of energy densities
as:
\begin{align}
  f_\text{mono} = f_\text{HS} - \frac{\alpha^\text{VDW}n}{kT}
\end{align}
with $\alpha^\text{VDW}$ given by
\begin{align}
  \alpha^\text{VDW} &= 2\pi\epsilon\int_\sigma^\infty r^2\phi(r)\,dr \\
  \intertext{using reduced units of $x = r/\sigma$, we have}
  &= 2\pi\sigma^3\int_1^\infty x^2\phi(x)\,dx \\
  &= 3b^\text{VDW}\epsilon\int_1^\infty x^2\phi(x)\,dx
\end{align}
where $b^\text{VDW}$ is the van der Waals size parameter. It corresponds to the volume excluded by two spherical particles of volume $b$: $b^\text{VDW} = 4b = 4\left(\pi\sigma^3/6\right)$.

@@

The pair potential $v(\1,\2)$ describes the interaction between two hard spheres. For the square-well model, the pair potential is
\begin{align}
  v_\text{SW}(\1,\2) &=
    \begin{cases}
      \infty & r < \sigma \\
      -\epsilon & \sigma < r < \lambda\sigma \\
      0 & \lambda\sigma < r
    \end{cases}
\end{align}
For this work I use $\lambda = 1.5$.

@@

Given the \emph{configuration integral} (the potenial energy component of the partition function)
\begin{align}
  Z_N &= \int d\mathbf{r}^N\, e^{-\beta V_N}
\end{align}
and the non-ideal part of the free energy (i.e.~the excess free energy) is
\begin{align}
  F &= -\kT\ln\left( \frac{Z_N}{V^N} \right)
\end{align}
where $V^N$ is the total volume \fixme{I think?}, then the derivative of the excess free energy with respect to the coupling parameter $\lambda$ is
\begin{align}
  \frac{\partial F}{\partial\lambda} &= -\kT\left( \frac{1}{Z_N}\frac{\partial Z_N}{\partial\lambda} \right) \\
  &= -\kT\frac{1}{Z_N}\int d\mathbf{r}^N\, e^{-\beta V_N}(-\beta)\frac{\partial V_N}{\partial\lambda} \\
  \hookrightarrow \beta\frac{\partial F}{\partial\lambda} &= \beta\left\langle V_N^{'}(\lambda) \right\rangle_\lambda \\
  \hookrightarrow \beta F &= \beta F_0 + \beta\int_0^1 d\lambda\, \left\langle V_N^{'}(\lambda) \right\rangle_\lambda
\end{align}
where $\left\langle V_N^{'}(\lambda) \right\rangle_\lambda$ is the canonical ensemble average. We can take a series expansion of this average:
\begin{align}\begin{split} % too long; use split environment, and align to a space (&{})
  \left\langle V_N^{'}(\lambda) \right\rangle_\lambda = &{} \left\langle V_N^{'}(\lambda)\right\rangle_{\lambda = 0} + (\lambda)\frac{\partial}{\partial\lambda}\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda}\bigg|_{\lambda = 0} \\ &{} + \mathcal{O}(\lambda^2)
\end{split}\end{align}
The derivative term is
\begin{align}
  \frac{\partial}{\partial\lambda}\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda} &= \left\langle V_N^{''} \right\rangle_\lambda - \beta\sigma_\lambda^2
\end{align}
where
\begin{align}
  \sigma_\lambda^2 &\equiv \left\langle \left[ V_N^{'}(\lambda) \right]^2 \right\rangle_\lambda - \left\langle V_N^{'}(\lambda) \right\rangle^2_\lambda
\end{align}
\fixme{I don't really understand how this derivative is \ldots derived}
Putting this together, we have
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta(1 - 0)\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}(1 - 0)^2\left(\left\langle V_N^{''}(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right)
  \end{split} \\
  \begin{split}
    =&{}  \beta F_0 + \beta\left\langle V_N^{'}(\lambda) \right\rangle_{\lambda = 0} \\ &{} - \beta\frac{1}{2}\left(\left\langle V_N^{''}(\lambda = 0) \right\rangle_{\lambda = 0} - \beta^2\sigma^2_{\lambda = 0} \right) \label{eqn:complicated-HTE}
  \end{split}
\end{align}
Now, we can define the total perturbation energy as
\begin{align}
  W_N &= \sum_{i=1}^N\sum_{j>i}^N w_\lambda(\mathbf{r}_i,\mathbf{r}_j)
\end{align}
Then, following from equation~\ref{eqn:small-perturbation},
\begin{align}
  V_N^{'} &= \frac{\partial}{\partial\lambda}\left(\sum_{i=1}^N\sum_{j>i}^N \lambda w(\mathbf{r}_i,\mathbf{r}_j)\right) \nonumber \\
  &= W_N \\
  \therefore V_N^{''} &= 0
\end{align}
With this in mind, we can simplify equation~\ref{eqn:complicated-HTE} to the \emph{high-temperature expansion}, first derived by Zwanzig:\cite{Zwanzig54}
\begin{align}
  \begin{split}
    \beta F = &{} \beta F_0 + \beta\left\langle W_N \right\rangle_0  \\ &{} - \frac{1}{2}\beta^2\left( \left\langle W_N^2 \right\rangle_0 - \left\langle W_N \right\rangle_0^2 \right) + \mathcal{O}(\beta^3)
  \end{split}
\end{align}
Zwanzig further showed that the $n^{th}$ term could be written in terms of the mean fluctuations $\left\langle \left[ W_N - \left\langle W_N\right\rangle_0 \right]^\nu \right\rangle_0$ with $\nu \leq n$.

Writing the statistical averages in terms of the pair density $n_\lambda^{(2)}(\1,\2)$ (i.e. the probability of having a sphere at \1 and \2 \fixme{make this more clear?}) gives us
\begin{align}
  \frac{\beta F}{N} &= \frac{\beta F_0}{N} + \frac{\beta}{2N}\int_0^1 d\lambda\, \iint d\1 d\2\, n_\lambda^{(2)}(\1,\2)w(\1,\2) \label{eqn:Henderson5-2-12}
\end{align}
The pair density itself can be expanded as
\begin{align}
  n_\lambda^{(2)}(\1,\2) &= n_0^{(2)}(\1,\2) + \lambda\frac{\partial n_\lambda^{(2)}(\1,\2)}{\partial\lambda}\bigg|_{\lambda = 0} + \mathcal{O}(\lambda^2)
\end{align}
Applying this to the \fixme{0th, 1st or 2nd, depending on how you think of it} term in equation~\ref{eqn:Henderson5-2-12}, we have
\begin{align}
  \frac{\beta F_1}{N} &= \frac{\beta}{2N}\iint d\1 d\2\, n_0^{(2)}(\1,\2)w(\1,\2) \\
  &= \frac{\beta n}{2}\int d\mathbf{r}_{12}\, g_0(\1,\2)w(\1,\2)
\end{align}
The second-order $\lambda$ term is quite a bit more complicated:
\begin{widetext}
 \begin{align}
   \begin{split}
   \frac{\beta F_2}{N} =&{} -\frac{\beta^2}{2}\left[ \frac{n}{2}\int d\2\, g_0(\1,\2)w^2(\1,\2) \right. \\ &{} + n^2 \iint d\2 d\3\, g_0^{(3)}(\1,\2,\3)w(\1,\2)w(\1,\3) \\ &{} \left. + \frac{n^3}{4}\iiint d\2 d\3 d\4\, \left( g_0^{(4)}(\1,\2,\3,\4) - g_0^{(2)}(\1,\2)g_0^{(2)}(\3,\4) \right)w(\1,\2)2(\3,\4) \right] \\ &{} - \frac{1}{4}S_0(0)\left[ \frac{\partial}{\partial n}\left( n^2\int d\2\, g_0(\1,\2)w(\1,\2) \right)^2 \right]
   \end{split}
 \end{align}
\end{widetext}
This term requires use to know the form of higher-order correlation functions and the structure function $S_0(k)$ of the reference system. These things are very difficult. Hence, we use approximations, as outlined in section~\ref{sub2sec:disp}.

@@

\subsection{Free Energies}\label{subsec:free-energies}
The $a_j(n)$ terms require even more simplification when ``converting'' from the mathematical formulation to Python code, to rid ourselves of troublesome integrals. Gil-Villegas gives us
\begin{align}
  a_1^\text{SW}(n) &= a_1^\text{VDW}(n)g^\text{HS}(1;\eta_{eff}) \\
  &= -4\eta\epsilon(\lambda^3 - 1)\frac{1 - \left( \eta_{eff}/2 \right)}{(1 - \eta_{eff})^3}
\end{align}
and
\begin{align}
  \eta_{eff} &= c_1\eta + c_2\eta^2 + c_3\eta^3
\end{align}
with
\begin{align}
  \left( \begin{array}{c}
    c_1 \\
    c_2 \\
    c_3
    \end{array} \right)
  &= \left( \begin{array}{ccc}
    2.25855 & -1.50349 & 0.249434 \\
    -0.669270 & 1.40049 & -0.827739 \\
    10.1576 & -15.0427 & 5.30827
    \end{array} \right)
  \left( \begin{array}{c}
    1 \\
    \lambda \\
    \lambda^2
    \end{array} \right)
\end{align}

\subsection{RG partition functions}\label{subsec:fbar-ubar}
I use the midpoint method for evaluating the integrals of $\bar{f}_D$ and $\bar{u}_D$.

@@

\begin{figure}
  \begin{center}
  \includegraphics[width=\columnwidth]{figs/phi-RG-lowT}
  \end{center}
  \caption{Free energy density for RG at low temperature}
  \label{fig:phi-RG-lowT}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width=\columnwidth]{figs/phi-RG-highT}
  \end{center}
  \caption{Free energy density for RG near critical temp}
  \label{fig:phi-RG-highT}
\end{figure}

Figures~\ref{fig:phi-RG-lowT} and~\ref{fig:phi-RG-highT} show the grand free energy per unit volume at a low temperature, and near the critical point. It is clear that the program is not de-bugged.
